Topological Equivariant Artist Model for Visualization Library Architecture

### 2022-06-15 paper/main.tex
cleaning up why trivial/non-trivial bundles matter

@@

There is an assumption in visualization that the connectivity of the elements in a dataset are preserved in the visual representation; roughly topology provides a method for encoding this connectivity \cite{wilkinsonGrammarGraphics2005}. Bertin gives the examples of discrete data values should be represented by discrete entities such as points, quantitive continuous data by lines, and surfaces through area marks \cite{bertinSemiologyGraphicsDiagrams2011a}. In \autoref{fig:related-work:continuity:ktypes} the records of the GHCN\cite{lawrimoreGlobalHistoricalClimatology2011} weather station table can be considered disconnected from each other and therefore have a 0D continuity. This means each distinct row can be reduced to a discrete point and it is expected  that each rows will be encoded as a discrete visual element, such as a standalone scatter point. Each station can also be considered a time series of weather observation since temperature and pressure are sampled from a 1D continuous measurement space. The connectivity of these observations can be modeled as an interval and the data is encoded as a line since that visual element is also 1D continuous. The measurements at set of stations on any given day is a sparse sampling of a 2D geospatial grid and the connectivity can be modeled as a 2D plane. Instead of interpolation, which is how continuity was preserved in the timeseries plot, in this instance the connectivity is preserved by plotting the stations in the map such that their relative distances are preserved. Topological continuity allows us to encode the connectivity of the data in a uniform manner such that we can then generalize the preservation of that continuity in the visual representation beyond points, lines, areas, and other case by case continuities.

@@

\subsection{Uniform Abstract Data Representation}
\label{sec:atct:fiber-bundles}
We model data using a mathematical representation of data that can encode topological properties, field types, and data values in a uniform. This mathematical representation is called a fiber bundle and its use as an abstract data representation in visualization was first proposed by Butler\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}. A fiber bundle $(\dtotalc, \dbasec, \pi, \dfiberc)$ is a structure with topological spaces $\dtotalc, \dfiberc, \dbasec$ and continuous surjective map $\pi: \dtotalc \rightarrow \dbasec$\cite{FiberBundle2020}.

@@

The \textcolor{base}{base space} is a topological space \dbasec\ with points $\dbasepoint \in \dbase$ and topology $\mathcal{T}_{\dbase}$. For our purposes, $\dbase$ is a set of points that encodes the underlying connectivity of the dataset. For example, $\dbase$ could be the discrete points, interval, or plane in \autoref{fig:related-work:continuity:ktypes}. On this set \dbase, we can define subsets of connected elements, called opensets \openset. An openset \openset\ is a generalization of an open interval, meaning an openset contains the points inside a bounded region on X and excludes the points on the boundary. A topology $\mathcal{T}_{\dbase}$ defined on \dbase\ is a collection of opensets such that the empty set and \dbase are in $\mathcal{T}_{\dbase}$,and every union and intersection of opensets in $\mathcal{T}_{\dbase}$ are also in $\mathcal{T}_{\dbase}$ \cite{TopologyVsTopology,bradleyTopologyCategoricalApproach2020}. The topology is encoding the notion that the connectivity is expected to be preserved regardless of how small an openset or large a union of opensets.

The \textcolor{fiber}{fiber space} is a topological space \dfiberc\ that is the preimage of the projection function $\pi$ at a point in the base space $\dfiber_{\dbasepoint} = \pi^{-1}(\dbasepoint)$, meaning that all points in a fiber over a space $\dfiber_{\dbasepoint}$ map to the same point $\dbasepoint$. Fibers of a bundle are isomorphic $\dfiberc \simeq \dfiberc_{\dbasepointc}\;\forall \dbasepointc \in \dbase$, which means that the space of possible values is the same set of values over all points in the base space.  The fiber space can be used to encode the space of all possible values of a data field, which is analogous to a CS data type, and the fibers can be named, as described by Spivak \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES}.

The \textcolor{section}{sections} $\dsection$ of a bundle are continuous functions that bind the base space with a set of values in the total space $\dtotal$ and $\textcolor{set}{\Gamma}$ denotes the set of all sections in a bundle over an openset.

\begin{equation}
  \label{eq:atct:fb:sections}
  \cgamma{\opensetc}{\dtotalc\restriction_{\opensetc}} \coloneqq \big\{\dsectionc: \opensetc\rightarrow \dtotalc\restriction_{\opensetc} \; \bigm{\vert} \pi(\dsectionc(\dbasepointc)) = \dbasepointc\;for\, all\; \dbasepointc \in \opensetc \big\} 
\end{equation}
Each section function takes as input a point in the basespace $\dbasepoint$ and returns a single point in the fiber. Extending Spivak's fiber bundle description of a table, we define the values returned by a section as named and typed records. Since the fiber space is a topological space, it does not really have restrictions on the dimension of the types it can encode or the structure of those types. For example, wind has a speed component and a direction component and we can consider those separately $\dfiber_{speed} = \mathbb{R}, \dfiber_{direction} = \mathbb{R}$ or as one fiber$\dfiber_{wind} = \mathbb{R}^{2}$.

The base space and the section evaluated on the base space form a sort of key-value pair, as described by Munzner \cite{munznerVisualizationAnalysisDesign2014}. She imbues semantic meaning to the topology, wherein the key is an independent attribute used as a look up key to retrieve the value in the dataset. She defines the value as a dependent attribute. In this work we use a more general notion of the topological space as an indexing space for the data records. Therefore, we do not impose any semantic constraints on what the fiber and base space encode. For example, a data cube of temperature can be encoded such that every fiber is a set of 2D temperature surfaces $\dfiber_{tempsurf}$, the base space is a 1D continuous interval, and each section returns a surface $\dsection(\dbasepoint) = f(x,y)$ where x and y are already parameterized. Alternatively, the fiber can encode each cell of the data cube $\dfiber_{lat} \times \dfiber_{lon} \times \dfiber_{temp}$, the base space could be a 3D volume, and each section returns a $\dsection(\dbasepoint) = \{lat, lon, temp\}$ record.

Fiber bundles where the fibers are not identical are called non-trivial bundles. Non-trivial bundles allow us to generalize bundles to many different types of total spaces. We can also decompose non-trivial bundles into trivial bundles, which are bundles where all the fibers are identical. This is because all fiber bundles are locally trivial, which means that for a small enough subset of the data $\opensetc \subseteq \dbasec$, all the fibers are equivalent $\pi^{-1}(\opensetc)\xrightarrow{\equivc} \opensetc \times \dfiberc$. A bundle may be globally trivial, meaning that $\dtotalc = \dbasec \times \dfiberc$, and generally can be thought of as a twisted product of total and base spaces \cite{munkresElementsAlgebraicTopology1984}. Using trivial and locally trivial bundles allows us to encode the possible data records in a bundle as a single fiber $\dfiber$ type rather than having to maintain a fiber type for each point \dbasepoint.

@@

@@

\subsection{Mathematical Structure of Data}
\label{sec:atct:sheaves:measurments}
We define structure on data as the allowable \textcolor{action}{transformations on the bundle}, which we formalize as actions $\dfuncc$ on sections of that sheaf. Actions  $act: X \times Y \tightarrow Y$ are elements of one set $X$ applied to elements of another set $Y$ such that the action in X applied to Y returns an element in Y\cite{nlab:action}. Actions are a generalization of applying a function to an element and returning an element of the same type. Our sheaf action  $\dfunc$ is a generalization of the classification of measurement scales by the actions allowed on the data values\cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale, thomasMathematizationNotMeasurement2014}, which often serves as the basis of the equivariance described in \autoref{sec:related-work:equivariance}.

We separate data transformations into two types, transformations on the base space $(\dfunchc, \dfuncpullc)$ and transformations on the fiber space $\dfunctc$. These diagrams are seperated because the base space transformation and the pullback are functors on the openset object and associated set of sections, while the fiber transformation takes a single fiber to a different fiber in the set of sections.

>>>>

\note{Spelling out vis library component behavior w/ alg top + cat theory lets us be super general in terms of inputs/outputs (what data, what vis) while super specific about the constraints the components need to satisfy to transform that input to that output in a structure preseving way}

@@

There is an assumption in visualization that the connectivity of the elements in a dataset are preserved in the visual representation; roughly topology provides a method for encoding this connectivity \cite{wilkinsonGrammarGraphics2005}. Bertin gives the examples of discrete data values should be represented by discrete entities such as points, quantitive continuous data by lines, and surfaces through area marks \cite{bertinSemiologyGraphicsDiagrams2011a}. In \autoref{fig:related-work:continuity:ktypes} the records of the GHCN\cite{lawrimoreGlobalHistoricalClimatology2011} weather station table can be considered disconnected from each other and therefore have a 0D continuity. This means each distinct row can be reduced to a discrete point and it is expected  that each rows will be encoded as a discrete visual element, such as a standalone scatter point. Each station can also be considered a time series of weather observation since temperature and pressure are sampled from a 1D continuous measurement space. The connectivity of these observations can be modeled as an interval and the data is encoded as a line since that visual element is also 1D continuous. The measurements at set of stations on any given day is a sparse sampling of a 2D geospatial grid and the connectivity can be modeled as a 2D plane. Instead of interpolation, which is how continuity was preserved in the timeseries plot, in this instance the connectivity is preserved by plotting the stations in the map such that their relative distances are preserved. Encoding connectivity using the formalism of topological spaces allows us to encode the connectivity of the data in a uniform manner such that we can then generalize the preservation of that connectivity in the visual representation rather than constructing a condition specifically for each type of topology.

@@

\subsection{Uniform Abstract Data Representation}
\label{sec:atct:fiber-bundles}
We model data using a mathematical representation of data that can encode topological properties, field types, and data values in a uniform manner using a structure from algebraic topology called a fiber bundle. We extend Butler's proposal of bundles as abstract visualization data type \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989} by incorporating Spivak's methodology for encoding named data types from his fiber bundle representation of relational databases \cite{spivakDatabasesAreCategories2010, spivakSIMPLICIALDATABASES}. We build on this work to describe how to encode the connectivity of the data as a topological space, separately encode the fields as their own topological space with a typing system, and express the mappings between these two spaces.

A fiber bundle $(\dtotalc, \dbasec, \pi, \dfiberc)$ is a structure with topological spaces $\dtotalc, \dfiberc, \dbasec$ and continuous surjective map $\pi: \dtotalc \rightarrow \dbasec$\cite{FiberBundle2020}.

@@

\subsubsection{Connectivity: Base Space \dbase}
The \textcolor{base}{base space} is a topological space \dbasec\ with points $\dbasepoint \in \dbase$ and topology $\mathcal{T}_{\dbase}$. For our purposes, $\dbase$ is a set of points that encodes the underlying connectivity of the dataset. For example, $\dbase$ could be the discrete points, interval, or plane in \autoref{fig:related-work:continuity:ktypes}. On this set \dbase, we can define subsets of connected elements, called opensets \openset. An openset \openset\ is a generalization of an open interval, meaning an openset contains points inside a bounded region on \dbase and excludes the points on the boundary. A topology $\mathcal{T}_{\dbase}$ defined on \dbase\ is a collection of opensets such that the empty set and \dbase are in $\mathcal{T}_{\dbase}$,and every union and intersection of opensets in $\mathcal{T}_{\dbase}$ are also in $\mathcal{T}_{\dbase}$ \cite{TopologyVsTopology,bradleyTopologyCategoricalApproach2020}. The topology is encoding the notion that the connectivity is expected to be preserved regardless of how small an openset or large a union of opensets.

\subsection{Field Types: Fiber Space \dfiber}
The \textcolor{fiber}{fiber space} is a topological space \dfiberc\ that is the preimage of the projection function $\pi$ at a point in the base space $\dfiber_{\dbasepoint} = \pi^{-1}(\dbasepoint)$, meaning that all points in a fiber over a space $\dfiber_{\dbasepoint}$ map to the same point $\dbasepoint$. The fiber space can be used to encode the space of all possible values of a data field and the fibers can be named, as described by Spivak \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES}. The space of all possible values is analogous to a CS data type; therefore the fiber space can be used to encode the data types of the data fields. This set theoretic formulation of field types also supports encoding valid operations on the type as part of the type, akin to methods on an object, as part of the field structure. For example, the Steven's measurement types are defined by the mathematical group structure on the measurment fields \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale, thomasMathematizationNotMeasurement2014}, and often serve as the basis of the equivariance described in \autoref{sec:related-work:equivariance}.

\subsection{Data: Section \dsection}
The \textcolor{section}{sections} $\dsection$ of a bundle are continuous functions that bind the base space with a set of values in the total space $\dtotal$ and $\textcolor{set}{\Gamma}$ denotes the set of all sections in a bundle over an openset.

\begin{equation}
  \label{eq:atct:fb:sections}
  \cgamma{\opensetc}{\dtotalc\restriction_{\opensetc}} \coloneqq \big\{\dsectionc: \opensetc\rightarrow \dtotalc\restriction_{\opensetc} \; \bigm{\vert} \pi(\dsectionc(\dbasepointc)) = \dbasepointc\;for\, all\; \dbasepointc \in \opensetc \big\} 
\end{equation}
Each section function takes as input a point in the basespace  and returns a single point in the fiber  $\dsection: \dbasepoint \rightarrow \dfiber_{\dbasepoint}$. In a trivial bundle $\dtotalc = \dbasec \times \dfiberc$ such that all fibers are equal $\dfiberc \equiv \dfiberc_{\dbasepointc}\;\forall \dbasepointc \in \dbase$; therefore we can define a global section $\dsection \in \Gamma(\dbase, \dfiber)$. In non-trivial (general) bundles, fibers are isomorphic $\dfiberc \simeq \dfiberc_{\dbasepointc}\;\forall \dbasepointc \in \dbase$ but not necessarily equal, which means that we usually cannot define a global section. Instead, all fiber bundles are locally trivializable, $\dtotal\restriction_{\openset_{k}} = \openset_{k}\times \dfiber$ for all open neighborhoods $\dbasepoint \openset \dbase$ for all points $\dbasepoint \in \dbase$. This allows us to define sections locally, meaning $\dsection\restriction_{\openset} \in \Gamma(\openset_{\dbasepoint}, \dtotal\restriction_{\openset_{\dbasepoint}})$. All data can be expressed as a collection of local sections, which are maps from the indexing space \dbase\ to the field space \dbase, and data that is embedded in trivial bundles can be expressed as a single section.

@@

\subsection{intrepreation}
Extending Spivak's fiber bundle description of a table, we define the values returned by a section as named and typed records. Since the fiber space is a topological space, it does not really have restrictions on the dimension of the types it can encode or the structure of those types. For example, wind has a speed component and a direction component and we can consider those separately $\dfiber_{speed} = \mathbb{R}, \dfiber_{direction} = \mathbb{R}$ or as one fiber$\dfiber_{wind} = \mathbb{R}^{2}$. The base space and the section evaluated on the base space form a sort of key-value pair, as described by Munzner \cite{munznerVisualizationAnalysisDesign2014}. She imbues semantic meaning to the topology, wherein the key is an independent attribute used as a look up key to retrieve the value in the dataset. She defines the value as a dependent attribute. In this work we use a more general notion of the topological space as an indexing space for the data records. Therefore, we do not impose any semantic constraints on what the fiber and base space encode. For example, a data cube of temperature can be encoded such that every fiber is a set of 2D temperature surfaces $\dfiber_{tempsurf}$, the base space is a 1D continuous interval, and each section returns a surface $\dsection(\dbasepoint) = f(x,y)$ where x and y are already parameterized. Alternatively, the fiber can encode each cell of the data cube $\dfiber_{lat} \times \dfiber_{lon} \times \dfiber_{temp}$, the base space could be a 3D volume, and each section returns a $\dsection(\dbasepoint) = \{lat, lon, temp\}$ record.

@@

\subsection{Mathematical Structure of Data}
\label{sec:atct:sheaves:measurments}
Equivariance in a visual context was formally defined as the preservation of a binary operator from one domain to another by Mackinlay\cite{mackinlayAutomaticDesignGraphical1987}. We generalize binary operations to actions $\dfunc \in \dfuncset$ on elements of the bundle $\delement \in \dtotal$. This is expressed with a composition function  $act: \dfuncset \times E \tightarrow \dtotal$ that applies the action to the element such that $\dfunc(\delement_0) = \delement_1, \delement_0, \delement_1 \in \dfunc$. We generalize to a family of actions because that allows for expanding the set of allowable transformations on the data beyond a single operator.

We separate data transformations into two types, transformations on the base space $(\dfunchc, \dfuncpullc)$ and transformations on the fiber space $\dfunctc$. These diagrams are separated because the base space transformation and the pullback are functors on the openset object and associated set of sections, while the fiber transformation takes a single fiber to a different fiber in the set of sections.
